Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-18 main.tex
Replace \natp by \nat in main.tex

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size

\newcommand{\MAPm}{\hat \mu_n}
\newcommand{\MAPt}{\hat \natp_n}

@@

\paragraph{Motivation.} 
We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.
We want the simplest one.
We hope  to get a new result by combining tools from statistics and optimization. 
\paragraph{Exponential Family.}
The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model 
\begin{equation}
	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{equation}
where $\logpart$ is the log-partition function (aka normalization factor) 
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
\end{align}

\begin{example}
	These frames contain the trailing example of this paper: a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
	The density of a centered normal variable is
\begin{align}
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other
\begin{align}
	\natp = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
	= x^2 \natp - \logpart(\natp) \\
	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
\end{align}
\end{example}

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
The second identity entails that $\logpart$ is strictly-convex. 
At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
\end{align}
It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
\begin{figure}[ht]
	\centering
	\includegraphics[width= .6\textwidth]{figs/duality} \\
	\includegraphics[width=.7\textwidth]{figs/2d/mirrormap.pdf}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.Top figure reproduced from \citet{kunstner2020homeomorphic}. Bottom figure represents $\cN(\mu, \sigma^2)$.}
	\label{fig:duality}
\end{figure}
\begin{example}
For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
\begin{align}
	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
\end{align}
which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.
\end{example}

\paragraph{Conjugate Prior}
One conjugate prior for $p(X|\natp)$ is
\begin{align}
    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp ; \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
\end{align}
with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
We can re-write this prior as 
\begin{align}
    p(\natp) \propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \; ,
\end{align}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.

\begin{example}
In the case of $\cN(0,\sigma^2)$, the logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
In particular,
\begin{align}
	p(\natp) 
	&\propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \\
    &\propto \exp( \half[n_0] \log(-\natp) + n_0\mu_0 \natp ) \\
	&\propto (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
\end{align}
from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\natp \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$. After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.
\end{example}

\paragraph{Maximum A Posteriori (MAP).}
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
\begin{align}
    -\log p(\cD|\natp)p(\natp) 
    = (n_0+n) \logpart (\natp) 
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
\end{align}
Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
\end{align}
such that the MAP is
\begin{align}
    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
\end{align}
The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.

\section{Open Problem}
For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{align}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\natp^*) ; p(.|\natp)) 
	= \bregman (\natp ; \natp^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\natp ; \natp^*)
    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}
\begin{important}
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
	\label{eq:bregmanMAP}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
\end{align}
where the outer expectation is on the dataset $X_1, \dots, X_n$?
\end{important}

@@

\begin{example}
	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
\end{example}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} \leq \log(z) 
	\label{eq:log_bound} 
	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

@@

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{equation}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{n_0 +n}{a+ (n-2)_+}
		\label{eq:theta_expectation_bound}
	\end{equation}
	where $(x)_+ = \max(0,x)$.
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Let us  define
 \begin{align}
	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
 \end{align}
The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.

@@

\section{Euclidean Behaviour}
Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space. 
\TODO{Add base measure to add here a discussion of the euclidean case -- the mean of a gaussian}
\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
    \leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz.
Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.
\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.

@@

\subsection{Primal Expectation Pivot}
Let $\tilde \theta_n := \expect{\hat \theta_n}$ and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the primal and dual parameters of the primal expectation of the MAP.
As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
	\expect{\bregman(\MAPt; \natp^*)} 
	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
	\label{eq:primal_pivot}
\end{align}

>>>>

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\nat}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size

\newcommand{\MAPm}{\hat \mu_n}
\newcommand{\MAPt}{\hat \nat_n}

@@

\paragraph{Motivation.} 
We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.
We want the simplest one.
We hope  to get a new result by combining tools from statistics and optimization. 
\paragraph{Exponential Family.}
The exponential family member with sufficient statistic $T$ and natural parameter $\nat$ is the model 
\begin{equation}
	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
\end{equation}
where $\logpart$ is the log-partition function (aka normalization factor) 
\begin{align}
    \logpart(\nat) = \log \int e^{\nat^\top T(x)} dx \; .
\end{align}

\begin{example}
	These frames contain the trailing example of this paper: a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
	The density of a centered normal variable is
\begin{align}
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other
\begin{align}
	\nat = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
	= x^2 \nat - \logpart(\nat) \\
	\implies \logpart (\nat) = - \half \log(-\nat)  + \half \log(\pi) 
\end{align}
\end{example}

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p(X|\nat)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\nat) &= \Cov_\nat[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\nat$ and mean parameters $\mu$.
The second identity entails that $\logpart$ is strictly-convex. 
At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\mu) = \langle \mu, \nat \rangle - \logpart(\nat) \; .
\end{align}
It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
\begin{figure}[ht]
	\centering
	\includegraphics[width= .6\textwidth]{figs/duality} \\
	\includegraphics[width=.7\textwidth]{figs/2d/mirrormap.pdf}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\nat, \meanp$.Top figure reproduced from \citet{kunstner2020homeomorphic}. Bottom figure represents $\cN(\mu, \sigma^2)$.}
	\label{fig:duality}
\end{figure}
\begin{example}
For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \nat - \logpart(\nat)$ to get the entropy
\begin{align}
	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
We can also take gradient and derivative of $\logpart(\nat)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
\begin{align}
	\nabla\logpart(\nat) &= \frac{-1}{2 \nat} = \sigma^2 = \mu = \E[X^2] \\
	\nabla^2\logpart(\nat) &= \frac{1}{2 \nat^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
\end{align}
which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.
\end{example}

\paragraph{Conjugate Prior}
One conjugate prior for $p(X|\nat)$ is
\begin{align}
    p(\nat) &\propto \exp( - n_0 \bregman(\nat ; \nat_0) )
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior, and $\bregman(\nat ; \nat_0)$ is the Bregman divergence induced by $\logpart$ between $\nat$ and $\nat_0$
\begin{align}
    \bregman (\nat ; \nat_0)
    & = \logpart(\nat) - \logpart(\nat_0) 
    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle
\end{align}
with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. 
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\nat_0$.
We can re-write this prior as 
\begin{align}
    p(\nat) \propto 
    \exp( -n_0 \logpart (\nat) 
    + \langle n_0 \mu_0, \nat \rangle ) \; ,
\end{align}
which is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.

\begin{example}
In the case of $\cN(0,\sigma^2)$, the logpartition function is $\logpart(\nat) = -\log(-\nat) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
In particular,
\begin{align}
	p(\nat) 
	&\propto 
    \exp( -n_0 \logpart (\nat) 
    + \langle n_0 \mu_0, \nat \rangle ) \\
    &\propto \exp( \half[n_0] \log(-\nat) + n_0\mu_0 \nat ) \\
	&\propto (-\nat)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\nat)} / Z
\end{align}
from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\nat \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$. After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.
\end{example}

\paragraph{Maximum A Posteriori (MAP).}
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\nat) = n_0 (\logpart(\nat)  - \nat^\top \meanp_0 ) + \cst
\end{align*}
Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\nat)$ is
\begin{align}
    -\log p(\cD|\nat)p(\nat) 
    = (n_0+n) \logpart (\nat) 
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
\end{align}
Minimizing this expression over $\nat$ yields the Maximum A Posteriori estimate
\begin{align}
    \hat \nat = \argmin_\nat -\log p(\cD|\nat) + n_0 \bregman(\nat ; \nat_0)
\end{align}
such that the MAP is
\begin{align}
    \nabla \logpart(\hat \nat_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
\end{align}
The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

\section{Open Problem}
For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[X\sim p(.|\nat^*)]{-\log p(X|\nat) + \log p(X|\nat^*) }
	= \KL( p(.|\nat^*) ; p(.|\nat)) \; .
\end{align}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\nat^*) ; p(.|\nat)) 
	= \bregman (\nat ; \nat^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\nat ; \nat^*)
    = \logpart(\nat) - \langle \nat , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\nat^*) ; p(.|\nat))
    = \bregman (\nat ; \nat^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}
\begin{important}
The question is: how does this quantity behave when $\nat$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[X_i\sim \nat^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
	\label{eq:bregmanMAP}
	\expect[X_i\sim \nat^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
\end{align}
where the outer expectation is on the dataset $X_1, \dots, X_n$?
\end{important}

@@

\begin{example}
	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \nat_n; \nat_* ) 
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
\end{example}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \nat_n; \nat_* )   = \phi(\frac{ \nat_n}{\nat_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} \leq \log(z) 
	\label{eq:log_bound} 
	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

@@

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{equation}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		\leq \frac{n_0 +n}{a+ (n-2)_+}
		\label{eq:theta_expectation_bound}
	\end{equation}
	where $(x)_+ = \max(0,x)$.
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Let us  define
 \begin{align}
	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
 \end{align}
The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.

@@

\section{Euclidean Behaviour}
Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space. 
\TODO{Add base measure to add here a discussion of the euclidean case -- the mean of a gaussian}
\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq L \norm{\mu^* - \mu} + \norm{\nat} \norm{\mu^* - \mu}
    \leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz.
Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.
\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.

@@

\subsection{Primal Expectation Pivot}
Let $\tilde \theta_n := \expect{\hat \theta_n}$ and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the primal and dual parameters of the primal expectation of the MAP.
As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
\begin{align}
	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
	\expect{\bregman(\MAPt; \nat^*)} 
	&= \bias{\bregman(\E[\MAPt]; \nat^*)}
	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
	\label{eq:primal_pivot}
\end{align}
